Counting And Confusion: Bayesian Rate Estimation With Multiple Populations

### 2013-01-15 many.tex
Edits to V and VI.

Suppose we attempt to detect gravitational wave signals in a data
stream by matched filtering in the frequency domain against a set of
$N$ template waveforms \citep[e.g.,][]{findchirppaper,LVC2011}.  In
our simplistic model, we suppose the data stream consists of
stationary Gaussian noise with a power spectral density $S(f)$
combined additively with some number of gravitational wave signals.
We assume that the signals are sufficiently rare that they do not
overlap in the data stream.  The signal to noise ratio (SNR) of a
template, $h(f)$, given data, $d(f)$, is
\begin{equation}
  \rho_h \equiv \frac{\left\langle h, d \right\rangle}{\sqrt{\left
      \langle h, h \right\rangle}},
\end{equation}
where $\left \langle \cdot \right\rangle$ denotes the noise-weighted
inner product:
\begin{equation}
  \left\langle a, b \right\rangle \equiv 4 \Re \int_0^\infty df\,
  \frac{a^*(f) b(f)}{S(f)}.
\end{equation}
We suppose for simplicity that the templates are sufficiently distinct
that
\begin{equation}
  \left\langle h_i, h_j \right\rangle \simeq \delta_{ij}.
\end{equation}
In the following subsection, we will generalize the model to
overlapping templates.  We rank candidate events by their maximum SNR
over the entire template bank,
\begin{equation}
  x \equiv \max_{h} \rho_h,
\end{equation}
and consider only events that have a maximum SNR above some threshold,
$x > \xmin$.

@@

The SNR of a gravitational wave signal in an interferometric detector
scales as $1/d$ \citep{Finn1992}, where $d$ is the distance to the
source.  Ignoring cosmological effects, the number of sources scales
as $d^3$.  Thus, we expect that the foreground cumulative distribution
of events will follow
\begin{equation}
  \label{eq:analytic-foreground-rate}
  \hat{F}(x) = 1 - \frac{\xmin^3}{x^3}.
\end{equation}

@@

In Figure \ref{fig:analytic-rate-recovery}, we show the marginalized
posterior densities for the foreground and background rates.  (Refer
to Eq.~\eqref{eq:rate-shape-posterior}.)  Figure
\ref{fig:analytic-rate-foreground-probs} shows the posterior
foreground probability for each event marginalized over all other
events' types and the foreground and background rates.

We can compare these results to results obtained using the two
approximations described earlier, the loudest-event statistic and the
foreground-dominated statistic. The marginalized distribution for the
foreground rate using these alternatives are shown in
Figure~\ref{fig:altmeth}. In this case, the loudest event had
$x_N\simeq 9.47$. The loudest-event statistic depends on a
specification of the maximum, $R_{\rm max}$, for the background
rate. We show results for $R_{\rm max}=\infty$, i.e., the improper
prior, and $R_{\rm max}=10000$. The results for other reasonable
choices of $R_{\rm max} = 100, 1000, 100000$ etc. gave exactly the
same posterior, since $\hat{b}(x_N) R_{\rm max} \ll \hat{f}(x_N)$ for
all these choices and we are therefore in the regime where the
posterior is insensitive to $R_{\rm max}$. To apply the
foreground-dominated statistic we must specify a threshold above which
we assume all events are foreground. It is reasonable to do this based
on a specification for the relative probability of an event being
fore/background, $\hat{f}(x)/\hat{b}(x)=p_{\rm thresh}$. Setting
$p_{\rm thresh}=0.99$ gives $x_{\rm min}=4.07$ and there are $N=18$
(11 foreground and 7 background) events exceeding that
threshold. Setting $p_{\rm thresh}=0.5$ gives $x_{\rm min}=3.82$ and
there are $N=30$ (11 foreground and 19 background) events exceeding
that threshold.  Each of these thresholds gives a biased estimate of
the rate because there are background events still above threshold.
The ``omniscient'' threshold of $x_\mathrm{min} = 4.38$ produces $N=7$
(7 foreground and 0 background) events in this data set, and therefore
an unbiased estimate, but of course this threshold can only be
determined because we can examine the synthetic foreground and
background data samples.  We show results for the first two choices of
$x_{\rm min}$ in Figure~\ref{fig:altmeth}; the omniscient choice
produces essentially the same posterior as our full analysis.

@@

In the context of gravitational wave experiments, additional information on the ranking statistic distributions for the foreground and background can be obtained using injections into and timeslides of the data respectively. This information can be readily incorporated in the current framework by assuming there is another set of $N_I$ events with ranking statistics $\{ w_i \}$, known to be drawn from the foreground distribution ($f_i=1$) and a set of $N_T$ events with ranking statistics $\{z_i\}$ known to be drawn from the background distribution ($f_i=0$). These events will typically not be drawn with the correct rate parameters, so they do not contribute to the estimates of $R_f$ and $R_b$, but they do contribute an extra factor
\be
\prod_{l=1}^{N_I} \hat{f}(w_l, \theta ) \prod_{m=1}^{N_T} \hat{b}(z_i, \theta) 
\ee
to the right hand sides of Eqs~(\ref{eq:posterior}) and (\ref{eq:rate-shape-posterior}). This approach provides a way to incorporate extra information into the analysis in order to simultaneously fit for the shape of the background and foreground as well as the rates. In the limit that there are many more events in the timeslide and injection data set, this will reduce to the analysis that was described above with fixed ranking statistic distributions $\hat{f}(x)$ and $\hat{b}(x)$ given by the injection and time slide data. We note that this analysis makes the assumption that the background distribution is the same in the time slide and real data and that the foreground distribution is the same between the injection and real data. The former assumption is probably reasonable, modulo non-stationarities in the detector data, but the latter relies on knowledge of the relative the astrophysical rates of different events, which is more uncertain. These astrophysical uncertainties could be handled with a hybrid approach, in which injections are used to characterize the statistic distribution for sources of a particular type, while additional rate or shape parameters are introduced to characterize the variation in the astrophysical rate of mergers as a function of source type.

@@

In this paper, we have developed a Bayesian framework for rate
estimation when the data consists of a mixture of foreground and
background events.  We demonstrated the application of this framework
using several examples from gravitational-wave data analysis in the
presence of signatures of binary mergers and noise triggers, and
astronomical image analysis in the presence of several populations of
stars.  We showed that this framework is generally superior to both
the loudest-event statistic and and the foreground-dominated
statistic.

Throughout this paper, we have assumed that the shape of the
foreground and background distributions is known, or at least can be
modeled with several additional parameters.  This is not necessarily
easy to do.  For example, in the case of gravitational-wave data
analysis, the shape of the foreground distribution of events may
depend on the details of a complex data-analysis pipeline as well as
the astrophysical source distribution, while the background event
distribution depends on data quality and may deviate significantly
from the simple Gaussian-noise behavior modeled in section
\ref{sec:GW-example}.  Several approaches have been developed to
accurately model both distributions, e.g., through the use of injected
signals to model the foreground distribution \cite{ihope,CannonHannaKeppel:2012}.
However, this is a difficult problem (e.g., because of the need to
estimate the background at the very tails of the distribution), and
will require significant future work.

@@

\begin{acknowledgments}
  We thank Kipp Cannon, Chad Hanna, Drew Keppel, Richard
  O'Shaughnessy, and Ted von Hippel for discussions and suggestions
  about this manuscript.  IM and WMF acknowledge the hospitality of KITP, supported in part by the National Science Foundation under No. NSF Grant PHY11-25915.  CC's work was carried out  at the Jet Propulsion Laboratory, California Institute of Technology, under contract to the National Aeronautics and Space Administration.  CC also gratefully acknowledges support from NSF grant PHY1068881.  JG's work is supported by the Royal Society.  Copyright 2012.  All rights reserved.
\end{acknowledgments}

>>>>

Suppose we attempt to detect gravitational wave signals in a data
stream by matched filtering in the frequency domain against a set of
$N$ template waveforms \citep[e.g.,][]{findchirppaper,LVC2011}.  In
our simplistic model, we suppose the data stream consists of
stationary Gaussian noise with a power spectral density $S(f)$
combined additively with some number of gravitational wave signals.
We assume that the signals are sufficiently rare that they do not
overlap in the data stream.  The signal-to-noise ratio (SNR) of a
template, $h(f)$, given data, $d(f)$, is
\begin{equation}
  \rho_h \equiv \frac{\left\langle h, d \right\rangle}{\sqrt{\left
      \langle h, h \right\rangle}},
\end{equation}
where $\left \langle \cdot \right\rangle$ denotes the noise-weighted
inner product:
\begin{equation}
  \left\langle a, b \right\rangle \equiv 4 \Re \int_0^\infty df\,
  \frac{a^*(f) b(f)}{S(f)}.
\end{equation}
We suppose for simplicity that the templates are sufficiently distinct
that
\begin{equation}
  \left\langle h_i, h_j \right\rangle \simeq \delta_{ij}.
\end{equation}
In the following subsection, we will generalize the model to
overlapping templates.  We rank candidate events by their maximum SNR
over the entire template bank,
\begin{equation}
  x \equiv \max_{h} \rho_h,
\end{equation}
and consider only events that have a maximum SNR above some threshold,
$x > \xmin$.

@@

The SNR of a gravitational-wave signal in an interferometric detector
scales as $1/d$ \citep{Finn1992}, where $d$ is the distance to the
source.  Ignoring cosmological effects, the number of sources scales
as $d^3$.  Thus, we expect that the foreground cumulative distribution
of events will follow
\begin{equation}
  \label{eq:analytic-foreground-rate}
  \hat{F}(x) = 1 - \frac{\xmin^3}{x^3}.
\end{equation}

@@

In Figure \ref{fig:analytic-rate-recovery}, we show the marginalized
posterior densities for the foreground and background rates (see
Eq.~\eqref{eq:rate-shape-posterior}).  Figure
\ref{fig:analytic-rate-foreground-probs} shows the posterior
foreground probability for each event marginalized over all other
events' types and the foreground and background rates.

We can compare these results to results obtained using the two
approximations described earlier, the loudest-event statistic and the
foreground-dominated statistic. The marginalized distribution for the
foreground rate using these alternatives are shown in
Figure~\ref{fig:altmeth}. In this case, the loudest event had
$x_N\simeq 9.47$. The loudest-event statistic depends on a
specification of the maximum, $R_{\rm max}$, for the background
rate. We show results for $R_{\rm max}=\infty$, i.e., the improper
prior, and $R_{\rm max}=10000$. The results for other reasonable
choices of $R_{\rm max} = 100, 1000, 100000$ etc. gave exactly the
same posterior, since $\hat{b}(x_N) R_{\rm max} \ll \hat{f}(x_N)$ for
all these choices and we are therefore in the regime where the
posterior is insensitive to $R_{\rm max}$. To apply the
foreground-dominated statistic we must specify a threshold above which
we assume all events are foreground. It is reasonable to do this based
on a specification for the relative probability of an event being
fore/background, $\hat{f}(x)/\hat{b}(x)=p_{\rm thresh}$. Setting
$p_{\rm thresh}=0.99$ gives $x_{\rm min}=4.07$ and there are $N=18$
(11 foreground and 7 background) events exceeding that
threshold. Setting $p_{\rm thresh}=0.5$ gives $x_{\rm min}=3.82$ and
there are $N=30$ (11 foreground and 19 background) events exceeding
that threshold.  Each of these thresholds gives a biased estimate of
the rate because there are background events still above threshold.
The ``omniscient'' threshold of $x_\mathrm{min} = 4.38$ produces $N=7$
(7 foreground and 0 background) events in this data set, and therefore
an unbiased estimate, but of course this threshold can only be
determined because we can examine the synthetic foreground and
background data samples. The threshold may seem obvious from a 
visual examination of Figure \ref{fig:analytic-rate-foreground-probs}; 
however, the construction of this figure relies on the application of the full
framework in the first place.  We show results for the first two choices of
$x_{\rm min}$ in Figure~\ref{fig:altmeth}; the omniscient choice
produces essentially the same posterior as our full analysis.

@@

In the context of gravitational wave experiments, additional information on the ranking statistic distributions for the foreground can be obtained using mock signal injections into the data, while distributions for the background can be estimated by analyzing time slides of data sets from different detectors relative to each other \cite[e.g.,][]{ihope}. This information can be readily incorporated in the current framework by assuming there is another set of $N_I$ events with ranking statistics $\{ w_i \}$, known to be drawn from the foreground distribution ($f_i=1$) and a set of $N_T$ events with ranking statistics $\{z_i\}$ known to be drawn from the background distribution ($f_i=0$). These events will typically not be drawn with the correct rate parameters, so they do not contribute to the estimates of $R_f$ and $R_b$, but they do contribute an extra factor
\be
\prod_{l=1}^{N_I} \hat{f}(w_l, \theta ) \prod_{m=1}^{N_T} \hat{b}(z_i, \theta) 
\ee
to the right hand sides of Eqs~(\ref{eq:posterior}) and (\ref{eq:rate-shape-posterior}). This approach provides a way to incorporate extra information into the analysis in order to simultaneously fit for the shape of the background and foreground as well as the rates. In the limit that there are many more events in the timeslide and injection data set, this will reduce to the analysis that was described above with fixed ranking statistic distributions $\hat{f}(x)$ and $\hat{b}(x)$ given by the injection and time slide data. We note that this analysis makes the assumption that the background distribution is the same in the time slide and real data and that the foreground distribution is the same between the injection and real data. The former assumption is probably reasonable, modulo correlations of non-gravitational-wave origin between data in different detectors, but the latter relies on knowledge of the relative the astrophysical rates of different events, which is more uncertain. These astrophysical uncertainties could be handled with a hybrid approach, in which injections are used to characterize the statistic distribution for sources of a particular type, while additional rate or shape parameters are introduced to characterize the variation in the astrophysical rate of mergers as a function of source type.

@@

In this paper, we have developed a Bayesian framework for rate
estimation when the data consists of a mixture of foreground and
background events.  We demonstrated the application of this framework
using several examples from gravitational-wave data analysis in the
presence of signatures of binary mergers and noise triggers, and
astronomical image analysis in the presence of several populations of
stars.  We showed that this framework is generally superior to both
the loudest-event statistic and the foreground-dominated
statistic.

Through most of this paper, we have assumed that the shape of the
foreground and background distributions is known, or at least can be
modeled with several additional parameters.  This is not necessarily
easy to do.  For example, in the case of gravitational-wave data
analysis, the shape of the foreground distribution of events may
depend on the details of a complex data-analysis pipeline as well as
the astrophysical source distribution, while the background event
distribution depends on data quality and may deviate significantly
from the simple Gaussian-noise behavior modeled in section
\ref{sec:GW-example}.  Several approaches have been developed to
accurately model both distributions, e.g., through the use of injected
signals to model the foreground distribution \cite{ihope,CannonHannaKeppel:2012}.
However, this is a difficult problem (e.g., because of the need to
estimate the background at the very tails of the distribution), and
will require significant future work.  In Section \ref{sec:uncertain-dist}, 
we discussed some of the possible
approaches when the shapes of the background and foreground distributions
cannot be confidently described by models with a few adjustable parameters.

@@

\begin{acknowledgments}
  We thank Kipp Cannon, Chad Hanna, Drew Keppel, Richard
  O'Shaughnessy, and Ted von Hippel for discussions and suggestions
  about this manuscript.  IM and WMF acknowledge the hospitality of KITP, supported in part by the National Science Foundation under No. NSF Grant PHY11-25915.  CC's work was carried out  at the Jet Propulsion Laboratory, California Institute of Technology, under contract to the National Aeronautics and Space Administration.  CC also gratefully acknowledges support from NSF grant PHY1068881.  JG's work is supported by the Royal Society.  Copyright 2013.  All rights reserved.
\end{acknowledgments}
